---
title: "A4: Incarceration"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
library(leaflet)
library(readr)
library(tidyr)
library(maps)
library(mapdata)
```


### Introduction

The data tool (trends.vera.org) is used for collecting the population of prison 
and jail at county levels to better understand the current situation in the U.S. 
The problem domain is an analysis of incarceration trends in the U.S. I have 
chosen the dataset 
[Dataset about prison population per county/state](https://github.com/melaniewalsh/Neat-Datasets/blob/main/us-prison-pop.csv?raw=true). 
This dataset provides comprehensive data on the prison population, including 
variables such as total prison population, prison population by race (Black, 
White, Latinx, etc.), and prison population by gender. These variables are 
crucial for understanding the disparities and trends in incarceration across 
different demographic groups. Analyzing this data will allow us to explore how 
incarceration rates vary by race and gender, and how these trends have evolved 
over time, providing insights into systemic issues and areas that require policy
attention. The inclusion of detailed demographic data makes this dataset a 
valuable resource for identifying patterns of inequality and informing potential
reforms in the criminal justice system.




### Summary Information

```{r}
# Load datasets
prison_pop <- read_csv("https://raw.githubusercontent.com/melaniewalsh/Neat-Datasets/main/us-prison-pop.csv")
```

```{r}
colnames(prison_pop)
```

```{r}
# 1. Average total prison population across all counties
average_total_prison_pop <- prison_pop %>%
  summarise(avg_total_prison_pop = mean(total_prison_pop, na.rm = TRUE))

# 2. Highest Black prison population
highest_black_prison_pop <- prison_pop %>%
  filter(black_prison_pop == max(black_prison_pop, na.rm = TRUE)) %>%
  select(black_prison_pop)

# 3.Lowest White prison population
lowest_white_prison_pop <- prison_pop %>%
  filter(white_prison_pop == min(white_prison_pop, na.rm = TRUE)) %>%
  select(white_prison_pop)

# 4. Average of change in the total prison population over the past 10 years
change_over_time <- prison_pop %>%
  group_by(year) %>%
  summarise(avg_prison_pop = mean(total_prison_pop, na.rm = TRUE)) %>%
  mutate(change = avg_prison_pop - lag(avg_prison_pop)) 

recent_change <- change_over_time %>%
  filter(year %in% (max(year) - 9):max(year))

# Calculate the average change over the past 10 years
average_recent_change <- recent_change %>%
  summarise(avg_change = mean(change, na.rm = TRUE))



# 5. Average female prison population
average_female_prison_pop <- prison_pop %>%
  summarise(avg_female_prison_pop = mean(female_prison_pop, na.rm = TRUE))

# Print the summary values
average_total_prison_pop
highest_black_prison_pop
lowest_white_prison_pop
average_recent_change
average_female_prison_pop
```

The analysis of the incarceration data reveals several key insights.  The average 
total prison population across all counties is approximately 361.36. The highest 
Black prison population is 24,818.  Conversely, the lowest White prison population 
is 0.  Over the past 10 years, the average change in the total prison population 
has been approximately -2.41, indicating a slight decrease overall.  Additionally,
the average female prison population across all counties stands at approximately
31.93.  These values underscore the significant racial disparities in 
incarceration rates, highlighting the need for policy interventions to address 
these imbalances.  The data shows that Black populations are disproportionately 
incarcerated, and the decrease in the overall prison population suggests changes 
in incarceration policies or practices that may warrant further investigation.



### The Dataset

Who collected the data?  

The data was compiled through the collaboration of several organizations,
including the Census of Jails, the Census of State and Federal Adult Correctional 
Facilities, the FBI Uniform Crime Reporting



How was the data collected or generated?  

The data was consolidated through the cooperation of multiple organizations, 
and they supplemented it with additional information when there were gaps.



Why was the data collected?  

The data was collected for understand and know the incarceration trends at
county-levels in the U.S. comprehensively. Thus, we can a broad range of
understanding in jail and prison population but lack the specific and exact data. 



How many observations (rows) are in your data? 
```{r}
p_row <- nrow(prison_pop)
p_row
```
There are 153811 observations in my data.



How many features (columns) are in the data?  
```{r}
p_col <- ncol(prison_pop)
p_col
```


What, if any, ethical questions or questions of power do you need to consider 
when working with this data? 

Working with this prison population dataset requires careful consideration of 
ethical issues and power dynamics. There is a risk of reinforcing negative 
stereotypes about racial groups due to the disparities highlighted in the data.
Systemic biases in data collection, like over-policing and discriminatory 
sentencing, can skew results. Ensuring data privacy is crucial, and the impact 
of incomplete or missing data must be acknowledged. Ethical use of the data 
involves advocating for fair policies and accurate, transparent communication to 
avoid misinterpretation and promote informed discussions on criminal justice reform.



What are possible limitations or problems with this data?   (at least 200 words)

The dataset about prison populations per county/state from the Vera Institute of 
Justice, while comprehensive, has several limitations and potential problems. 
One major limitation is the presence of missing or incomplete data. Not all 
correctional facilities report data consistently, leading to gaps that can affect 
the reliability and accuracy of the analysis. This inconsistency can also result 
in significant variations in how different states and counties report their 
incarceration statistics, introducing potential biases and inaccuracies. 
Additionally, the dataset is historical and may not fully capture recent changes
in incarceration policies or practices. As criminal justice reform efforts gain
traction, changes in policies, such as the reduction of mandatory minimum sentences
or the decriminalization of certain offenses, may not be reflected in the data.
This lag can mislead analyses that aim to assess the current state of incarceration.
Another issue is the lack of qualitative data that can provide deeper insights into
the reasons behind the trends. Quantitative data alone cannot fully explain the 
socio-economic and political factors contributing to incarceration rates. Factors 
such as changes in law enforcement practices, economic conditions, and social
movements play a crucial role but are not captured in the dataset. Systemic biases
within the criminal justice system, such as over-policing in minority communities 
and discriminatory sentencing practices, can also influence the data. These biases 
can result in an over-representation of certain racial groups, particularly Black 
and Latinx populations, which can skew the analysis and potentially reinforce negative 
stereotypes. Furthermore, the dataset does not account for the lived experiences 
of incarcerated individuals or the broader impact of incarceration on communities. 
This lack of context can lead to an incomplete understanding of the issue and hinder
the development of holistic and effective policy solutions. Finally, the interpretation
of the data requires careful consideration. Statistical analyses can highlight 
disparities, but they do not always explain the underlying causes. Simplistic 
conclusions drawn from complex data can misinform policy decisions. It is essential
to approach the data with a critical and nuanced perspective, acknowledging its 
limitations and considering the broader socio-economic and political contexts that 
contribute to incarceration trends. In summary, while the dataset is a valuable 
resource for understanding incarceration trends in the U.S., its limitations and
potential biases must be carefully managed to ensure accurate and ethical use.



### Trends Over Time Chart

This chart shows the prison population trends for Black, White, and Latinx groups
from 1990 to the most recent year available in the dataset, illustrating how these 
populations have changed over time and highlighting significant disparities. The 
chart reveals a notable increase in both Black and White prison populations from 
1990 to around 2010, reflecting the impact of policies such as the War on Drugs. 
Around 2010, these populations begin to plateau and eventually decrease, suggesting 
the influence of criminal justice reforms aimed at reducing mass incarceration. 
Throughout the timeline, the Black prison population remains consistently higher 
than both White and Latinx populations, underscoring persistent racial disparities. 
The Latinx prison population shows a relatively stable trend with a slight increase 
until around 2005, followed by a decline, potentially reflecting different impacts 
of immigration policies and enforcement. Towards the most recent years, there is
a sudden drop in all three populations, which may be attributed to recent reforms,
data collection anomalies, or significant policy changes aimed at reducing prison 
populations. This visualization provides insights into racial disparities within 
the prison system and helps understand how various policies and societal changes 
have impacted these groups differently over time.




``````{r, warning = FALSE, message = FALSE}
# Prepare the data
trend_data <- prison_pop %>%
  filter(year >= 1990) %>%
  group_by(year) %>%
  summarise(
    black_prison_pop = sum(black_prison_pop, na.rm = TRUE),
    white_prison_pop = sum(white_prison_pop, na.rm = TRUE),
    latinx_prison_pop = sum(latinx_prison_pop, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(black_prison_pop, white_prison_pop, latinx_prison_pop), names_to = "race", values_to = "prison_pop")

# Create the line chart
ggplot(trend_data, aes(x = year, y = prison_pop, color = race)) +
  geom_line(size = 1) +
  labs(
    title = "Trends in Prison Population Over Time by Race",
    x = "Year",
    y = "Prison Population",
    color = "Race"
  )
```

### Variable Comparison Chart

The chart titled "Relationship Between Prison Population and Black Prison Population" 
is a scatter plot illustrating the correlation between the total prison population 
and the Black prison population across various counties. It shows a positive correlation, 
indicated by the upward-sloping blue regression line, suggesting that counties 
with higher total prison populations tend to have higher Black prison populations. 
The clustering of points around the line reflects a strong relationship, while 
outliers indicate counties with disproportionate Black prison populations. This
visualization was included to highlight potential racial disparities in incarceration
and to underscore areas where the Black prison population deviates from the overall 
trend, warranting further investigation and potential policy interventions.


```{r, warning = FALSE, message = FALSE}
prison_pop <- read_csv("https://raw.githubusercontent.com/melaniewalsh/Neat-Datasets/main/us-prison-pop.csv")
comparison_data <- prison_pop %>%
  filter(!is.na(total_prison_pop) & !is.na(black_prison_pop))

ggplot(data = comparison_data, aes(x = total_prison_pop, y = black_prison_pop)) +
  geom_point(color = "pink") +
  geom_smooth(method=lm, color= "blue",  se=FALSE) +
  labs(title = paste("Relationship Between Prison Population and Black",
                     "Prison Population"),
       x = "Total Prison Population",
       y = "Black Prison Population")
```


### Map

The map titled "Geographic Distribution of Total Prison Population by State" 
visualizes the total prison population across the United States using a color 
gradient from yellow to red, indicating lower to higher populations, respectively.
This map was included to provide a geographical perspective on the distribution 
of prison populations, highlighting regional disparities. The map reveals that 
states like California, Texas, and Florida have significantly higher prison populations,
as indicated by darker shades of red, while states in the Midwest and Northeast 
show lighter shades of yellow, indicating lower populations. This visualization 
underscores the regional variations in prison populations, which could be influenced
by state-specific policies, crime rates, and demographic factors. By illustrating 
these patterns, the map offers valuable insights for policymakers to understand 
where prison populations are concentrated and highlights areas that may benefit 
from targeted criminal justice reforms.


```{r}
state_abbrev_to_name <- data.frame(
  state = state.abb,
  full_name = tolower(state.name)
)

state_data <- prison_pop %>%
  group_by(state) %>%
  summarise(total_prison_pop = sum(total_prison_pop, na.rm = TRUE)) %>%
  left_join(state_abbrev_to_name, by = "state")

state_data <- state_data %>%
  rename(region = full_name)

us_states <- map_data("state")

map_data <- merge(us_states, state_data, by = "region", all.x = TRUE)

ggplot(map_data, aes(long, lat, group = group, fill = total_prison_pop)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(low = "yellow", high = "red", na.value = "grey50") +
  labs(
    title = "Geographic Distribution of Total Prison Population by State",
    fill = "Total Prison Population"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_text(hjust = 0.5)
  ) +
  coord_map()
```

